---
title: "Survival Analysis"
author: "Yusuf Tatlier"
---

## Survival Analysis

Survival analysis is a branch of statistics that is concerned with the time until a 
certain event (often some sort of failure event) occurs. It differs in this regard from 
time-independent statistical methods, like logistic regression, as it does not solely
look at whether an event occurs but also at the time. In this context, the survival function $s(t)$ specifies whether the event of interest has occured at time $t$.

## Kaplan-Meier estimator and curve

The Kaplan-Meier estimator provides a non-parametric estimate of the survival function.
Suppose we are analyzing a certain population, where the size of the (active) population at time $n$ will be denoted by $x_n$. If we denote the fraction of failures at time $n$ by $d_n=\frac{x_n-x_{n+1}}{x_n}$, the Kaplan-Meier estimator for time $T$ is given by:
$$ \prod_{n \leq T} (1-d_{n})$$
The Kaplan-Meier curve is basically a line plot of all the estimators that are obtained in this way. As the data in practice will be discrete, the curve will be flat between points.
Below we show an example.

```{r,warning=FALSE,error=FALSE,message=FALSE}
require(survival)

#Use the kidney dataset in the package survival
data(kidney)
surv_fit<-survfit(Surv(time, status) ~ 1, data = kidney)
#plot also shows the confidence interval
plot(surv_fit,main="Kaplan-Meier curve kidney dataset")

#It is also possible to compare the curve for different groups in the data separately using survfit
surv_fit_groups<-survfit(Surv(time, status) ~ sex, data = kidney)
plot(surv_fit_groups,col=2:3,main="Kaplan-Meier curves male/female groups kidney dataset")
legend("topright", legend = c("male","female"),text.col = 2:3)

#The plot indicates that the survival rates are different for men and women. We can also test
#this statistically
diff <- survdiff(Surv(time, status) ~ sex, data = lung)
pvalue=1 - pchisq(diff$chisq, length(diff$n) - 1)
#test shows a statistically significant difference
pvalue 
```

## Cox proportional hazards model

Given the hazard function $\lambda(t)$, which denotes the event rate at time $t$ conditional on survival until that time, the Cox proportional hazards model models the hazard function as follows:
$$  \lambda(t)=\lambda_0(t)exp(\beta_1x_1+...+\beta_nx_n) $$
Here $\lambda_0(t)$ is a base hazard function that is further scaled up or down per timestep depending on the covariates $x_1,...,x_n$ and coefficients $\beta_1,...,\beta_n$.
Note that if we have two different groups with two different hazard functions $\lambda_a(t)$ and $\lambda_b(t)$, the proportion of the two hazard functions will be time-independent.

```{r}
#par(mfrow=c(2,1))

#Univariate Cox model in which sex is used as covariate
cox_fit_uni<-coxph(Surv(time, status) ~ sex, data = kidney)
#plot also shows confidence interval
fit_sex<-cox.zph(cox_fit_uni)
#Note that the plot shows a clear slope, this is an indication that the sex covariate is statistically significant
plot(fit_sex,main="Univariate Cox model")

#Multivariate Cox model in which sex, age are used as covariate
cox_fit_mult<-coxph(Surv(time, status) ~ sex+age, data = kidney)
#plot also shows confidence interval
fit_age_sex<-cox.zph(cox_fit_mult)
plot(fit_age_sex,main="Multivariate Cox model")

#Note that the age covariate is not significant, this can also be seen from the plot as the plot is quite flat and doesn't indicate a significant slope.
print(cox_fit_mult)

#We can perform statistical tests in order to see whether the difference between survival curves of different groups is statistically significant.
#See below that for the covariate sex this is indeed the case. Note that a likelihood ratio
#test, Wald test and logrank test are given, all of being highly significant
summary(cox_fit_uni)
```

## Main References

* Survival analysis in R for Public Health, Imperial College London
* 'Survival' package in R (version 3.1-12), Terry M Therneau 